### 2022-02-15 chapters/onset/onset.tex
commentaires de Xaymara + qqs figures refaites
relecture II

\begin{abstract}
    For the last 8 years, Florida's Coral Reef (FCR) has suffered from severe coral loss caused by stony coral tissue loss disease (SCTLD). The outbreak reportedly initiated near Virginia Key in September 2014, during the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015. The disease then spread to the entirety of FCR, likely through waterborne transmission. Although the causative agent of SCTLD remains unknown, there is evidence that sediment can act as vector for the disease. Here we evaluate whether sediments produced or resuspended during dredging operations could have reached reefs where the disease was first reported and assess whether the timing of these sediment fluxes was consistent with the observed spread of the disease. We evaluated the transport of sediments caused by the expansion of PoM between November 2013 and September 2014 using a quasi-3D sediment transport model forced by currents from the high-resolution coastal ocean model SLIM. A particular attention was paid to the non conventional dredging operations performed without pumping the produced chopped rock particles. Our result suggest that the dredging had close to no direct impact on the coral reefs of Virginia Key. However, a significant fraction of the sediments produced reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites. Furthermore, using a biophysical disease agent transport model, we show that waterborne transmission from one of these sites to Virginia Key was possible before September 2014. This study brings new insight on the role that the deepening of PoM might have played in the onset of one of the worst coral outbreaks on record in the Caribbean.
\end{abstract}

>>>>

\begin{abstract}
    For the last 8 years, Florida's Coral Reef (FCR) has suffered from severe coral loss caused by stony coral tissue loss disease (SCTLD). The outbreak reportedly initiated near Virginia Key in September 2014, during the deepening of the Port of Miami (PoM) shipping channel, that took place between November 20, 2013 and March 16, 2015. The disease then spread to the entirety of FCR, likely through waterborne transmission. Although the causative agent of SCTLD remains unknown, there is evidence that sediments can act as vector for the disease. Here we evaluate whether sediments produced or resuspended during dredging operations could have reached reefs where the disease was first reported and assess whether the timing of these sediment fluxes was consistent with the observed spread of the disease. We evaluated the transport of sediments caused by the expansion of PoM between November 2013 and September 2014 using a quasi-3D sediment transport model forced by currents from the high-resolution coastal ocean model SLIM. A particular attention was paid to the non conventional dredging operations performed without pumping the produced chopped rock particles. Our results suggest that the dredging had close to no direct impact on the coral reefs of Virginia Key through sediment flux and sedimentation. However, a significant fraction of the sediments produced reached monitoring sites where the disease was reported prior to September 2014, suggesting that the dredging might have played a role in the onset of the outbreak at these sites. Furthermore, using a biophysical disease agent transport model, we show that waterborne transmission from one of these sites to Virginia Key was possible before September 2014. This study brings new insight on the role that the deepening of PoM might have played in the onset of one of the worst coral outbreaks on record in the Caribbean.
\end{abstract}

### 2022-02-15 chapters/sctld/article.tex
commentaires de Xaymara + qqs figures refaites
relecture II

These connectivity matrices are more easily handled by interpreting them as large graphs whose vertices are sub-reefs and whose edges represent connectivity pathways. They were analyzed using graph theory tools. In the present study, four potential connectivity measures were used to interpret the monthly computed graphs. These indicators are described in Table \ref{tab:indicator}. The first indicator is the weighted connectivity length (WCL), which gave the average dispersal distance from origin to destination for material produced on a given reef. The weighted connectivity of reef polygon $i$ is expressed as:
\begin{equation}
    \textrm{WCL}_i = \dfrac{\sum_j \tilde{C}_{ij} L_{ij}}{\sum_j \tilde{C}_{ij}}
\end{equation}
where $L_{ij}$ is the distance between origin reef $i$ and destination reef $j$. Another measure of the spreading potential of reef $j$ is its out-degree, \ie the product of the number of connections originating from reef $j$ by the quantity of disease agents it sent to the network. This indicator was obtained by computing the number of non-zero entries of row $i$ in the potential connectivity matrix $C$ and multiplying it with $\sum_j C_{ij}$. The information given by the out-degree was complemented by the fraction of disease agents produced on reef $i$ that successfully settled on a reef, called the fraction exchanged of reef $i$. This indicator is given by $\sum_{j} \tilde{C}_{ij}$. Finally, the isolation of reef $i$ in the network was given by its self-recruitment, \ie the fraction of disease agents settling on reef $i$ that originated from reef $i$, computed by $C_{ii}/\sum_jC_{ji}$. A large self-recruitment value indicates that little infectious material produced elsewhere settled on the reef and thus that it was isolated from the rest of the network. However, due to its relative nature, high self-recruitment could still be coupled with a large amount of exogenous diseased agents reaching the reef, although this quantity would be considered proportionally small in regard to the total mass of diseased material settling on the reef.

@@

The spread of the SCTLD throughout the FRT was simulated using a connectivity-based \cite{kermack1927contribution} SIR model. SIR models are among the most standard epidemiological models. They divide individuals into three compartments: susceptible ($S$), infectious ($I$) and removed ($R$). When affected by the disease, susceptible individuals become infectious and infect other susceptible individuals until they are removed from the system, either through recovery or death. Such models usually rely on the hypothesis of an homogeneous, well-mixed population. To account for the spatial heterogeneity of the FRT, the basic SIR formulation was modified by considering the fractions of susceptible ($S_j$), infectious ($I_j$) and removed ($R_j$) corals of each sub-reef $j$. Although the pathogen of SCTLD has not been identified, studies suggest a contagious mode of transmission \citep{aeby2019pathogenesis,muller2020spatial}. Here, we use 'infectious' to denote disease agents that could be passed from individual to individual, which are responsible for causing disease signs. We note that the disease agents could be biological or non-biological in nature. In the present study's epidemiological model, individual reefs interact through the exchange of disease agents as represented by the different connectivity matrices. For each sub-reef $j$ and at any time, the following relations hold: $0\leq S_j,I_j,R_j\leq 1$ and $S_j+I_j+R_j=1$. The evolution of these fractions through time is governed by the following equations:
\begin{equation}
    \begin{aligned}
        \dfrac{dS_j}{dt} &= -\beta\sum_i\dfrac{A_i}{A_j}I_i\tilde{C}_{ij}S_j - \beta'(I_j)S_jI_j \\
        \dfrac{dI_j}{dt} &= \beta\sum_i\dfrac{A_i}{A_j}I_i\tilde{C}_{ij}S_j + \beta'(I_j)S_jI_j - \sigma I_j \\
        \dfrac{dR_j}{dt} &= \sigma I_j
    \end{aligned}\label{eq:epidemio}
\end{equation}
where $\tilde{C}_{ij}$ is the entry corresponding to reef pair $(i,j)$ of the normalized potential connectivity matrix [-], $A_i$ is the area of reef polygon $i$ [km$^2$], $\sigma$ is the mortality rate [day$^{-1}$], and $\beta$ and $\beta'(I_j)$ are the inter- and intra-reef disease transmission rates [day$^{-1}$], respectively. In this model, diseased corals of sub-reef $i$ can 'infect' corals of sub-reef $j$ if there is non-zero probability of disease agents exchange from sub-reef $i$ to sub-reef $j$, given by $\tilde{C}_{ij}$. Moreover, to account for coral resistance to the disease, the intra-reef transmission function $\beta'(I_j)$ has the shape of a smooth step function of the fraction of infectious corals $I_j$ and is expressed as follows:
\begin{equation}
    \beta'(I_j) = \dfrac{\beta_0'}{2}(1+\tanh[(I_j-I_0)/\tau]),\label{eq:beta}
\end{equation}
where $I_0$ is a threshold on the infection population above which intra-reef transmission becomes significant, and $\tau$ is a measure of the interval over which the transition from low to high transmission occurs. As long as the fraction of infectious corals on sub-reef $j$ is below $I_0$, the only infection mechanism taking place is connectivity-driven transmission at rate $\beta$. Once the threshold is approached, intra-reef transmission with rate $\beta'_0$ is activated. A larger value of threshold $I_0$ corresponds to a greater resistance of corals to the disease, and therefore a slower spread of the disease within reef $j$. Coral reproduction and natural (\ie non SCTLD-related) death rates are not taken into account in this model, which amounts to assume that they balance each other. For this study the same values were used for $\beta$ and $\beta'_0$.
%This assumption seems reasonable since coral populations of the FRT were declining prior to the emergence of the SCTLD outbreak.

@@

In order to solve Eqs. \ref{eq:epidemio}, initial conditions were needed, \ie fractions of susceptible, infectious and recovered corals at the beginning of the simulated period. This information was constructed from 9 different field-collected datasets: (i) Coral Reef Evaluation and Monitoring Project (CREMP; 2014–2017), (ii) CREMP Presence/Absence Data (CREMP P\_A; 2016–2017), (iii) Southeast Florida Coral Reef Evaluation and Monitoring Project (SECREMP; 2014–2017), (iv) Florida Reef Resilience Program Disturbance Response Monitoring (FRRP; 2014–2017), (v) Hurricane Irma Rapid Reef Assessment (IRMA; 2017, \cite{viehman2018}), (vi) the Southeast Florida Action Network citizen science program (SEAFAN; 2014–2017), and (vii) the Southefrn Coral Disease Margin field effort (2017 and 2018; \cite{neely2018surveying}), (viii) Mote Marine Laboratory’s Field operations data (2018-2019) and (ix) data compiled through the citizen science BleachWatch program (2018). Every dataset provided data on the presence or absence of the SCTLD (or tissue loss consistent with the SCTLD case definition) within each survey. Some also provided detailed disease metrics such as the species affected and the disease prevalence, which was subsequently compiled into presence/absence of SCTLD data by surveyed site. The locations of these observations are shown in Fig. \ref{fig:stns}. Using this information, we first delineated the diseased zone by constructing the concave hull of the points where the disease was observed before May 2018. The reefs diseased prior to the beginning of our simulated period were then defined as the reefs located inside the constructed zone. The time of disease observation was then spatially interpolated on each reef of the diseased zone by kriging with a Gaussian semivariogram using Python \texttt{pyKrige} module \citep{murphy2014pykrige}. Assuming an initial state $(S,I,R)=(1-I_0, I_0, 0)$ when the disease was observed, the fractions of susceptible, infectious and removed corals on each reef of the diseased zone on the 1st May 2018 was finally approximated using Eqs. \ref{eq:simplified}. Reefs outside of the diseased zone were initialized with an entirely susceptible population.

@@

\begin{figure}
    \centering
    \includegraphics[width=.99\linewidth]{chapters/sctld/figures/hull_example_2.jpg}
    \caption{Method used to compute the disease front displacement during a simulated time interval.\textbf{a.} Concave hull $H_0$ of the diseased sub-reefs at the beginning of the simulated period. \textbf{b.} Concave hull $H_1$ of the sub-reefs diseased during the simulated time interval. \textbf{c.} Arrow showing the computed front displacement during simulated time interval between $H_0$ and $H_2$, the union of $H_0$ and $H_1$.}
    \label{fig:hull}
\end{figure}

@@

% 1. MODEL PARAMETERS
After calibration, we estimated the species-averaged basic reproduction number $\beta_0'/\sigma$ to be equal to $1.0835$. This value, being close to 1, suggests modeled diseased individuals were removed from the system almost as fast as susceptible individuals became diseased. This situation caused the fraction of infectious corals on the reefs to remain fairly low (\ie $\leq 0.4\%$) through time. These results suggest that only a small fraction of the colonies caused the disease to spread on the reef during the outbreak. The observation-based species-averaged transmission period of 6.45 days used in the model was a reasonable estimation of the disease transmission dynamics, being of the same order of magnitude as the experimentally-derived mean incubation period of 9.7 days. The difference between the two values can be explained by field measurement uncertainties as well as the inability to perfectly mimic field conditions in laboratory. In the present study, the same values were used for inter- and intra-reef 'infection' rates $\beta$ and $\beta_0'$, \ie disease agents 'infectiousness' was assumed to remain unchanged during their journey from reef to reef. This assumption is consistent with sensitivity analysis results, which showed that reducing the value of $\beta$ strongly impeded disease propagation through the network. This outcome suggests that, to reproduce the observed spread, inter- and intra-reef transmission rates must have similar magnitude, \ie that the causative agent is not degraded while traveling from reef to reef. Given that the etiology of SCTLD is still unknown, the lack of degradation of the SCTLD causative agent over distances of tens of kilometers between reefs may help to narrow the search.

@@

The fact that mean barotropic currents outperformed the other modes of transport can be explained by considering the trajectories of the particles used to model the transport of the disease causative agent. Due to the impact of winds on positively buoyant material, particles driven by surface currents are likely to be blown away from the reefs. Moreover, even when winds are pushing particles along the reef line, these particles spend less time over the same region than particles driven by mean and bottom currents. Smaller amounts of particle mass will therefore settle on reef polygons, leading to lower connection strengths in the potential connectivity matrix, \ie lower exchange of infectious material between reefs. Hence, despite being able to transport the disease over greater distances, surface currents are less efficient to drive the propagation of the disease. Particles driven by bottom currents, on the other hand, remain over the same region longer, producing larger entries of the potential connectivity matrix. Due to these large exchange probabilities between reefs, bottom currents are better at propagating the disease (Fig. \ref{fig:results}). Nevertheless, because bottom currents are slower, exchanges of disease agents occur within a more limited geographical range. Mean barotropic currents that carry particles greater distances while allowing for sufficiently large amounts of diseased material to settle on reef polygons, were thus best suited to propagate the disease at a speed similar to field-collected data (Fig. \ref{fig:results}).

>>>>

These connectivity matrices are more easily handled by interpreting them as large graphs whose vertices are sub-reefs and whose edges represent connectivity pathways. They were analyzed using graph theory tools. In the present study, four potential connectivity measures were used to interpret the monthly computed graphs. These indicators are described in Table \ref{tab:indicator}. The first indicator is the weighted connectivity length (WCL), which gave the average dispersal distance from origin to destination for material produced on a given reef. The weighted connectivity of reef polygon $i$ is expressed as:
\begin{equation}
    \textrm{WCL}_i = \dfrac{\sum_j \tilde{C}_{ij} L_{ij}}{\sum_j \tilde{C}_{ij}}
\end{equation}
where $L_{ij}$ is the distance between origin reef $i$ and destination reef $j$. Another measure of the spreading potential of reef $j$ is its out-degree, \ie~the product of the number of connections originating from reef $j$ by the quantity of disease agents it sent to the network. This indicator was obtained by computing the number of non-zero entries of row $i$ in the potential connectivity matrix $C$ and multiplying it with $\sum_j C_{ij}$. The information given by the out-degree was complemented by the fraction of disease agents produced on reef $i$ that successfully settled on a reef, called the fraction exchanged of reef $i$. This indicator is given by $\sum_{j} \tilde{C}_{ij}$. Finally, the isolation of reef $i$ in the network was given by its self-recruitment, \ie~the fraction of disease agents settling on reef $i$ that originated from reef $i$, computed by $C_{ii}/\sum_jC_{ji}$. A large self-recruitment value indicates that little infectious material produced elsewhere settled on the reef and thus that it was isolated from the rest of the network. However, due to its relative nature, high self-recruitment could still be coupled with a large amount of exogenous diseased agents reaching the reef, although this quantity would be considered proportionally small in regard to the total mass of diseased material settling on the reef.

@@

The spread of the SCTLD throughout the FRT was simulated using a connectivity-based \cite{kermack1927contribution} SIR model. SIR models are among the most standard epidemiological models. They divide individuals into three compartments: susceptible ($S$), infectious ($I$) and removed ($R$). When affected by the disease, susceptible individuals become infectious and infect other susceptible individuals until they are removed from the system, either through recovery or death. Such models usually rely on the hypothesis of an homogeneous, well-mixed population. To account for the spatial heterogeneity of the FRT, the basic SIR formulation was modified by considering the fractions of susceptible ($S_j$), infectious ($I_j$) and removed ($R_j$) corals of each sub-reef $j$. Although the pathogen of SCTLD has not been identified, studies suggest a contagious mode of transmission \citep{aeby2019pathogenesis,muller2020spatial}. Here, we use 'infectious' to denote disease agents that could be passed from individual to individual, which are responsible for causing disease signs. We note that the disease agents could be biological or non-biological in nature. In the present study's epidemiological model, individual reefs interact through the exchange of disease agents as represented by the different connectivity matrices. For each sub-reef $j$ and at any time, the following relations hold: $0\leq S_j,I_j,R_j\leq 1$ and $S_j+I_j+R_j=1$. The evolution of these fractions through time is governed by the following equations:
\begin{equation}
    \begin{aligned}
        \dfrac{dS_j}{dt} &= -\beta\sum_i\dfrac{A_i}{A_j}I_i\tilde{C}_{ij}S_j - \beta'(I_j)S_jI_j \\
        \dfrac{dI_j}{dt} &= \beta\sum_i\dfrac{A_i}{A_j}I_i\tilde{C}_{ij}S_j + \beta'(I_j)S_jI_j - \sigma I_j \\
        \dfrac{dR_j}{dt} &= \sigma I_j
    \end{aligned}\label{eq:epidemio}
\end{equation}
where $\tilde{C}_{ij}$ is the entry corresponding to reef pair $(i,j)$ of the normalized potential connectivity matrix [-], $A_i$ is the area of reef polygon $i$ [km$^2$], $\sigma$ is the mortality rate [day$^{-1}$], and $\beta$ and $\beta'(I_j)$ are the inter- and intra-reef disease transmission rates [day$^{-1}$], respectively. In this model, diseased corals of sub-reef $i$ can 'infect' corals of sub-reef $j$ if there is non-zero probability of disease agents exchange from sub-reef $i$ to sub-reef $j$, given by $\tilde{C}_{ij}$. Moreover, to account for coral resistance to the disease, the intra-reef transmission function $\beta'(I_j)$ has the shape of a smooth step function of the fraction of infectious corals $I_j$ and is expressed as follows:
\begin{equation}
    \beta'(I_j) = \dfrac{\beta_0'}{2}(1+\tanh[(I_j-I_0)/\tau]),\label{eq:beta}
\end{equation}
where $I_0$ is a threshold on the infection population above which intra-reef transmission becomes significant, and $\tau$ is a measure of the interval over which the transition from low to high transmission occurs. As long as the fraction of infectious corals on sub-reef $j$ is below $I_0$, the only infection mechanism taking place is connectivity-driven transmission at rate $\beta$. Once the threshold is approached, intra-reef transmission with rate $\beta'_0$ is activated. A larger value of threshold $I_0$ corresponds to a greater resistance of corals to the disease, and therefore a slower spread of the disease within reef $j$. Coral reproduction and natural (\ie~non SCTLD-related) death rates are not taken into account in this model, which amounts to assume that they balance each other. For this study the same values were used for $\beta$ and $\beta'_0$.
%This assumption seems reasonable since coral populations of the FRT were declining prior to the emergence of the SCTLD outbreak.

@@

In order to solve Eqs. \ref{eq:epidemio}, initial conditions were needed, \ie~fractions of susceptible, infectious and recovered corals at the beginning of the simulated period. This information was constructed from 9 different field-collected datasets: (i) Coral Reef Evaluation and Monitoring Project (CREMP; 2014–2017), (ii) CREMP Presence/Absence Data (CREMP P\_A; 2016–2017), (iii) Southeast Florida Coral Reef Evaluation and Monitoring Project (SECREMP; 2014–2017), (iv) Florida Reef Resilience Program Disturbance Response Monitoring (FRRP; 2014–2017), (v) Hurricane Irma Rapid Reef Assessment (IRMA; 2017, \cite{viehman2018}), (vi) the Southeast Florida Action Network citizen science program (SEAFAN; 2014–2017), and (vii) the Southefrn Coral Disease Margin field effort (2017 and 2018; \cite{neely2018surveying}), (viii) Mote Marine Laboratory’s Field operations data (2018-2019) and (ix) data compiled through the citizen science BleachWatch program (2018). Every dataset provided data on the presence or absence of the SCTLD (or tissue loss consistent with the SCTLD case definition) within each survey. Some also provided detailed disease metrics such as the species affected and the disease prevalence, which was subsequently compiled into presence/absence of SCTLD data by surveyed site. The locations of these observations are shown in Fig. \ref{fig:stns}. Using this information, we first delineated the diseased zone by constructing the concave hull of the points where the disease was observed before May 2018. The reefs diseased prior to the beginning of our simulated period were then defined as the reefs located inside the constructed zone. The time of disease observation was then spatially interpolated on each reef of the diseased zone by kriging with a Gaussian semivariogram using Python \texttt{pyKrige} module \citep{murphy2014pykrige}. Assuming an initial state $(S,I,R)=(1-I_0, I_0, 0)$ when the disease was observed, the fractions of susceptible, infectious and removed corals on each reef of the diseased zone on the 1st May 2018 was finally approximated using Eqs. \ref{eq:simplified}. Reefs outside of the diseased zone were initialized with an entirely susceptible population.

@@

\begin{figure}
    \centering
    \includegraphics[width=.99\linewidth]{chapters/sctld/figures/hull_example.png}
    \caption{Method used to compute the disease front displacement during a simulated time interval.\textbf{a.} Concave hull $H_0$ of the diseased sub-reefs at the beginning of the simulated period. \textbf{b.} Concave hull $H_1$ of the sub-reefs diseased during the simulated time interval. \textbf{c.} Arrow showing the computed front displacement during simulated time interval between $H_0$ and $H_2$, the union of $H_0$ and $H_1$.}
    \label{fig:hull}
\end{figure}

@@

% 1. MODEL PARAMETERS
After calibration, we estimated the species-averaged basic reproduction number $\beta_0'/\sigma$ to be equal to $1.0835$. This value, being close to 1, suggests modeled diseased individuals were removed from the system almost as fast as susceptible individuals became diseased. This situation caused the fraction of infectious corals on the reefs to remain fairly low (\ie~$\leq 0.4\%$) through time. These results suggest that only a small fraction of the colonies caused the disease to spread on the reef during the outbreak. The observation-based species-averaged transmission period of 6.45 days used in the model was a reasonable estimation of the disease transmission dynamics, being of the same order of magnitude as the experimentally-derived mean incubation period of 9.7 days. The difference between the two values can be explained by field measurement uncertainties as well as the inability to perfectly mimic field conditions in laboratory. In the present study, the same values were used for inter- and intra-reef 'infection' rates $\beta$ and $\beta_0'$, \ie~disease agents 'infectiousness' was assumed to remain unchanged during their journey from reef to reef. This assumption is consistent with sensitivity analysis results, which showed that reducing the value of $\beta$ strongly impeded disease propagation through the network. This outcome suggests that, to reproduce the observed spread, inter- and intra-reef transmission rates must have similar magnitude, \ie~that the causative agent is not degraded while traveling from reef to reef. Given that the etiology of SCTLD is still unknown, the lack of degradation of the SCTLD causative agent over distances of tens of kilometers between reefs may help to narrow the search.

@@

The fact that mean barotropic currents outperformed the other modes of transport can be explained by considering the trajectories of the particles used to model the transport of the disease causative agent. Due to the impact of winds on positively buoyant material, particles driven by surface currents are likely to be blown away from the reefs. Moreover, even when winds are pushing particles along the reef line, these particles spend less time over the same region than particles driven by mean and bottom currents. Smaller amounts of particle mass will therefore settle on reef polygons, leading to lower connection strengths in the potential connectivity matrix, \ie~lower exchange of infectious material between reefs. Hence, despite being able to transport the disease over greater distances, surface currents are less efficient to drive the propagation of the disease. Particles driven by bottom currents, on the other hand, remain over the same region longer, producing larger entries of the potential connectivity matrix. Due to these large exchange probabilities between reefs, bottom currents are better at propagating the disease (Fig. \ref{fig:results}). Nevertheless, because bottom currents are slower, exchanges of disease agents occur within a more limited geographical range. Mean barotropic currents that carry particles greater distances while allowing for sufficiently large amounts of diseased material to settle on reef polygons, were thus best suited to propagate the disease at a speed similar to field-collected data (Fig. \ref{fig:results}).

### 2022-02-15 chapters/sctld/sctld.tex
updated ref for onset
relecture II

\item \textbf{Dobbelaere, T.}, Muller, E. M., Gramer, L. J., Holstein, D. M., $\&$ Hanert, E. (2020). Coupled epidemio-hydrodynamic modeling to understand the spread of a deadly coral disease in Florida. \textit{Frontiers in Marine Science},1016. \href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.3389/fmars.2020.591881}.
\end{list}

\begin{abstract}
    For the last six years, the Florida Reef Tract (FRT) has been experiencing an outbreak of the Stony Coral Tissue Loss Disease (SCTLD). First reported off the coast of Miami-Dade County in 2014, the SCTLD has since spread throughout the entire FRT with the exception of the Dry Tortugas. However, the causative agent for this outbreak is currently unknown. Here we show how a high-resolution bio-physical model coupled with a modified patch Susceptible-Infectious-Removed epidemic model can characterize the potential causative agent(s) of the disease and its vector. In the present study, the agent is assumed to be transported within composite material (e.g., coral mucus, dying tissues, and/or resuspended sediments) driven by currents and potentially persisting in the water column for extended periods of time. In this framework, our simulations suggest that the SCTLD is likely to be propagated within neutrally buoyant material driven by mean barotropic currents. Calibration of our model parameters with field data shows that corals are diseased within a mean transmission time of 6.45 days, with a basic reproduction number slightly above 1. Furthermore, the propagation speed of the disease through the FRT is shown to occur for a well-defined range of values of a disease threshold, defined as the fraction of diseased corals that causes an exponential growth of the disease in the reef site. Our results present a new connectivity-based approach to understand the spread of the SCTLD through the FRT. Such a method can provide a valuable complement to field observations and lab experiments to support the management of the epidemic as well as the identification of its causative agent.
\end{abstract}

>>>>

\item \textbf{Dobbelaere, T.}, Muller, E. M., Gramer, L. J., Holstein, D. M., \& Hanert, E. (2020). Coupled epidemio-hydrodynamic modeling to understand the spread of a deadly coral disease in Florida. \textit{Frontiers in Marine Science}, 7, 1016. \href{https://www.sciencedirect.com/science/article/pii/S1463500322000026}{doi: 10.3389/fmars.2020.591881}.
\end{list}

\begin{abstract}
    For the last six years, the Florida Reef Tract (FRT) has been experiencing an outbreak of the Stony Coral Tissue Loss Disease (SCTLD). First reported off the coast of Miami-Dade County in 2014, the SCTLD has since spread throughout the entire FRT with the exception of the Dry Tortugas. However, the causative agent for this outbreak is currently unknown. Here we show how a high-resolution bio-physical model coupled with a modified patch Susceptible-Infectious-Removed epidemic model can characterize the potential causative agent(s) of the disease and its vector. In the present study, the agent is assumed to be transported within composite material (\textit{e.g.}~coral mucus, dying tissues, and/or resuspended sediments) driven by currents and potentially persisting in the water column for extended periods of time. In this framework, our simulations suggest that the SCTLD is likely to be propagated within neutrally buoyant material driven by mean barotropic currents. Calibration of our model parameters with field data shows that corals are diseased within a mean transmission time of 6.45 days, with a basic reproduction number slightly above 1. Furthermore, the propagation speed of the disease through the FRT is shown to occur for a well-defined range of values of a disease threshold, defined as the fraction of diseased corals that causes an exponential growth of the disease in the reef site. Our results present a new connectivity-based approach to understand the spread of the SCTLD through the FRT. Such a method can provide a valuable complement to field observations and lab experiments to support the management of the epidemic as well as the identification of its causative agent.
\end{abstract}

### 2022-02-15 include/abstract.tex
commentaires de Xaymara + qqs figures refaites
relecture II

Coral populations have declined dramatically worldwide under the combined effects of climate change and local anthropogenic stressors. In addition to ocean warming and acidification, threats to coral reefs in the Caribbean include frequent and virulent disease outbreaks, and intensifying hurricanes. In particular, Florida's Coral Reef (FCR) is currently facing a multi-year outbreak of the stony coral tissue loss disease (SCTLD). First observed in 2014 during the monitoring of the PortMiami Deep Dredge Project (PMDDP), the disease has now spread through the entire FCR and has been reported in numerous territories of the Caribbean. However, the propagation of the disease through FCR seemed to slow down when it reached its southwestern end, between the Marquesas and the Dry Tortugas (DRTO). Although the causative agent of the disease remains unknown, studies showed that its transmission was likely waterborne and that sediments could act as SCTLD vectors. The hydrodynamics should therefore be highly explanatory of its spread. Furthermore, Florida is a prime landfall target for hurricanes during the reproduction period of corals. In addition to causing wholesale destruction of the reefs, hurricanes might thus also impact larval dispersal through wind-wave-induced currents. Here, we use the high resolution coastal ocean model SLIM to capture transport processes in the FCR, and study the impact of SCTLD and hurricanes to coral reefs. First, a coupled epidemio-hydrodynamic model is developed to reproduce and understand the observed spread of SCTLD in Florida. A sediment transport model is then used to evaluate the impact of the PMDDP on the observed onset of the disease in 2014. Finally, a coupled wave-current model is implemented to study the impact of Hurricane Irma (2017) on transport processes over reefs in the Florida Keys. Assuming the dispersal of disease agents within neutrally buoyant material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling before reaching the DRTO to eddy activity in the Loop Current/Florida Current system. Furthermore, the results of our sediment transport model suggest that the PMDDP might have triggered the onset of the disease. Finally, our coupled wave-current model showed the important impact of wave-induced currents on transport processes over reefs during hurricanes. This thesis highlights the potential of models informed by and confronted against field knowledge as powerful tools to inform the management of complex marine ecosystems such as the FCR.

%The population of corals throughout the world has declined dramatically and many coral species are now listed as threatened Avoiding global warming being no longer an option, reef managers now focus on damage control and restoration strategies. The Florida Reef Tract, currently facing an unprecedented coral disease outbreak and where healthy corals grown in nurseries are replanted on damaged reefs, is a good application case of such reef management strategies. In order to maximize the efficiency of these efforts, a good knowledge of exchanges of infectious materials and larvae between reefs is required. However, connectivity patterns between reefs remain mostly unknown. To address this issue, we will use the high-resolution ocean model SLIM to simulate connectivity patterns between reefs and then identify reefs with the highest restoration potential as well the ones that are most at risk of being infected by the outbreak.

%The dispersal of coral larvae away from their natal habitats is an important process for coral reef ecosystems, but remains poorly understood and hard to gauge. This thesis presents a spatially explicit numerical modelling approach to simulate larval dispersal in Australia's Great Barrier Reef (GBR), and in doing so to estimate connectivity between the thousands of reefs that comprise it. A high resolution, finite element ocean model, SLIM, is used to build a biophysical model of larval dispersal capable of estimating larval exchange between reefs in large sections of the the GBR. Tools from network science are employed to investigate the community structure of the GBR's connectivity network, and these tools are used to reveal the presence of clusters of highly inter-connected reefs. Subsequently, the model is used to estimate the extent of connectivity between reefs of different morphologies and depths in the GBR. Finally, the possible impacts of climate change on connectivity in the GBR are assessed, and estimates are made of the likely changes to coral larval dispersal patterns over the coming century.

%Titan, Saturn’s largest moon, is a unique object in the Solar system as much as it has a substantial atmosphere, a surface with a complex interplay of geological processes and an outer ice shell overlyinga subsurface ocean. The climate on this icy satellite boasts a multi-phase hydrological cycle where methane plays a role similar to that of water on Earth. Surface lakes and seas filled with liquid methane and ethane are found in the polar regions. The subsurface ocean, on the other hand, is filled with liquid water. This liquid layer lying beneath Titan’s surface allows for the large surface deformation observed.

%The first objective of this thesis was to adapt an Earth-based geophysical and environmental model, SLIM (www.climate.be/slim), to Titan’s specific conditions in order to study the tidal motion in the surface lakes and seas. The modified model was applied to the largest lake in the southern hemisphere, Ontario Lacus, and the two largest seas, Kraken and Ligeia Maria, which are located in the northern hemisphere. The predicted surface elevation and velocity fields are part of the data needed to develop an exploration mission focusing on the surface lakes and seas of Titan. The normal modes of Ontario Lacus were also numerically studied. While resonantly forced modes could generate significant liquid motion, the natural periods are much shorter than the period of the astronomical forcings. Therefore, only atmospheric forcings could resonantly force the normal modes. Strong wind conditions corresponding to a stormy event are required to generate a surface elevation resulting in significant shoreline variations.

%Then, the model was used to study the tides of Titan’s global subsurface ocean. To this end, the shallow water equations and the 3D hydrostatic equations under the Boussinesq approximation were modified in order to take into account the ice shell lying at the top of the ocean. The method was adapted to solve the equations on a sphere. The effects of the shell are represented by adding a surface friction and a surface pressure term. The shell decreases the ocean surface elevation and slows down the flow without modifying significantly the global patterns of these fields. The magnitude of these variations depends on the mechanical behaviour of the ice shell. Surface and bottom heat fluxes play a significant role in the liquid motion of the ocean. The interactions between the tidal motion and the thermally driven flow resulting from the surface heat flux were studied by means of the 3D version of SLIM. The surface heat flux significantly impacts the velocity field, both in terms of magnitude and orientation while the influence on the ocean’s surface elevation is small. These results could be useful for astrobiologists paying attention to the ocean habitability and to validate hypotheses about Titan’s internal structure.

>>>>

Coral populations have declined dramatically worldwide under the combined effects of climate change and local anthropogenic stressors. In addition to ocean warming and acidification, threats to coral reefs in the Caribbean include frequent and virulent disease outbreaks, and intensifying hurricanes. In particular, Florida's Coral Reef (FCR) is currently facing a multi-year outbreak of the stony coral tissue loss disease (SCTLD). First observed in 2014 during the monitoring of the PortMiami Deep Dredge Project (PMDDP), the disease has now spread through the entire FCR and has been reported in numerous territories of the Caribbean. However, the propagation of the disease through FCR seemed to slow down when it reached its southwestern end, between the Marquesas and the Dry Tortugas (DRTO). Although the causative agent of the disease remains unknown, studies showed that its transmission was likely waterborne and that sediments could act as SCTLD vectors. The hydrodynamics should therefore be highly explanatory of its spread. Furthermore, Florida is a prime landfall target for hurricanes during the reproduction period of corals. In addition to causing wholesale destruction of the reefs, hurricanes might thus also impact larval dispersal through wind-wave-induced currents. Here, we use the high resolution coastal ocean model SLIM to capture transport processes in the FCR, and study the impact of SCTLD and hurricanes to coral reefs. First, a coupled epidemio-hydrodynamic model is developed to reproduce and understand the observed spread of SCTLD in Florida. A sediment transport model is then used to evaluate the impact of the PMDDP on the observed onset of the disease in 2014. Finally, a coupled wave-current model is implemented to study the impact of Hurricane Irma (2017) on transport processes over reefs in the Florida Keys. Assuming the dispersal of disease agents within neutrally buoyant material, our model successfully reproduced the observed spread of SCTLD through the FCR and linked its apparent stalling before reaching the DRTO to eddy activity in the Loop Current/Florida Current system. Furthermore, the results of our sediment transport model suggest that the PMDDP might have triggered the onset of the disease. Finally, our coupled wave-current model showed the important impact of wave-induced currents on transport processes over reefs during hurricanes. This thesis highlights the potential of models informed by and confronted against field knowledge as powerful tools to inform the management of complex marine ecosystems such as Florida's Coral Reef.
